---
title: "COVID 19 Data Analysis"
author: "G. Siegfried"
date: "2024-08-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Description


### Required Packages
```{r declaring}
# List of packages to check and load
packages <- c("ggplot2", "dplyr", "readr", "lubridate", "tidyr", "stringr")
```

#### Installing Packages
This function checks if the packages listed above are installed and installs any that are missing.
```{r setupPackages, include=FALSE}
# Function to check and install packages
check_and_install <- function(package) {
  if (!requireNamespace(package, quietly = TRUE)) {
    message(paste("Installing package:", package))
    install.packages(package)
  } else {
    message(paste("Package", package, "is already installed."))
  }
}
# Apply the function to each package
sapply(packages, check_and_install)
```

#### Loading Libraries:
This function will load the libraries of the required packages listed above.
```{r setupLibrary, include=FALSE}
# Function to check and load packages
check_and_load <- function(package) {
  if (!requireNamespace(package, quietly = TRUE)) {
    message(paste("The package", package, "is not installed. Installing now."))
    install.packages(package)
  }
  library(package, character.only = TRUE)
}
# Apply the function to each package
sapply(packages, check_and_load)
```

### Import Data:
```{r import-data}
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"
url_lookup <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"
file_names <- c("time_series_covid19_confirmed_US.csv", "time_series_covid19_confirmed_global.csv", "time_series_covid19_deaths_US.csv", "time_series_covid19_deaths_global.csv")
urls <- str_c(url_in,file_names)

global_cases <- read_csv(urls[2])
global_deaths <- read_csv(urls[4])
us_cases <- read_csv(urls[1])
us_deaths <- read_csv(urls[3])

uid <- read_csv(url_lookup)
```

### Transform & Clean Data:
```{r transform-data}
uid <- uid %>%
  select(-c(Lat,Long_, Combined_Key, code3, iso2, iso3, Admin2, UID, FIPS))

global_cases <- global_cases %>%
  pivot_longer(cols = -c('Province/State','Country/Region', Lat, Long),
               names_to = "date",
               values_to = "cases") %>%
  select(-c(Lat,Long))

global_deaths <- global_deaths %>%
  pivot_longer(cols = -c('Province/State','Country/Region', Lat, Long),
               names_to = "date",
               values_to = "deaths") %>%
  select(-c(Lat,Long))

global_combo <- global_cases %>%
  full_join(global_deaths) %>%
  rename(Country_Region = 'Country/Region',
         Province_State = 'Province/State') %>%
  mutate(date = mdy(date)) %>%
  unite("Combined_Key",
        c(Province_State, Country_Region),
        sep = ", ",
        na.rm = TRUE,
        remove = FALSE)

us_cases <- us_cases %>%
  pivot_longer(cols = -(UID:Combined_Key),
               names_to = "date",
               values_to = "cases") %>%
  select(Admin2:cases) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat,Long_))

us_deaths <- us_deaths %>%
  pivot_longer(cols = -(UID:Population),
               names_to = "date",
               values_to = "deaths") %>%
  select(Admin2:deaths) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat,Long_)) 

us_combo <- us_cases %>%
  full_join(us_deaths)

global <- global_combo %>%
  left_join(uid, by = c("Province_State", "Country_Region")) %>%
  select(Province_State, Country_Region, date, cases, deaths, Population, Combined_Key)
```

### Summary Data:

### Visualize Data:
```{r visualize-data}

US_by_State <- us_combo %>%
  group_by(Province_State, Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths), Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths *1000000 / Population) %>%
  select(Province_State, Country_Region, date, cases, deaths, deaths_per_mill, Population) %>%
  ungroup()

US_Totals <- US_by_State %>%
  group_by(Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths), Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths *1000000 / Population) %>%
  select(Country_Region, date, cases, deaths, deaths_per_mill, Population) %>%
  ungroup()

US_Totals %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = cases)) +
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) +
  scale_y_log10() +
  theme(legend.position="bottom", axis.text.x = element_text(angle = 90)) +
  labs(title = "COVID19 in US", y=NULL)

state <- "California"
US_by_State %>%
  filter(Province_State == state) %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = cases)) +
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) +
  scale_y_log10() +
  theme(legend.position="bottom", axis.text.x = element_text(angle = 90)) +
  labs(title = str_c("COVID19 in ", state), y=NULL)
```

### Model Data
```{r model_data}

mod <- lm(deaths ~ cases, data = US_Totals)
summary(mod)

US_Totals_w_Pred <- US_Totals %>% mutate(pred = predict(mod))
US_Totals_w_Pred %>% ggplot() +
  geom_point(aes(x = cases, y = deaths), color = "blue") +
  geom_point(aes(x = cases, y = pred), color = "red")

```

## Conclusion:

There is a lot of bias within this data due to the errors in collecting the data. I have done my best to represent it as true as possible.